---
title: "Covid-19 in Ireland"
output: 
  html_document:
    theme: united
params:
  output_dir: "docs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

We perform simple mathematical modelling from government-published data, of recorded cases of Covid-19 in the Republic of Ireland.

For a refresher on exponential growth, please see the 3Blue1Brown [video](https://www.youtube.com/watch?v=Kas0tIxDvrg).

## The Data

Data is published daily at an Irish Government [website](https://www.gov.ie/en/news/7e0924-latest-updates-on-covid-19-coronavirus/).

```{r message=FALSE}
d <- read.csv(file="~/src/covid/data.csv", header=TRUE, sep=",")
d[is.na(d)] <- 0
d$date <- as.Date(d$date)
d$day <- as.integer(d$date - d$date[1])
d$total_cases <- cumsum(d$cases)
d$total_deaths <- cumsum(d$deaths)
```

## Exponential Phase

From the plot below, it is clear that as of 19 March, we are currently in the exponential-growth phase. (An r-squared value of 0.97 indicates a very good fit.)

```{r}
f <- log(total_cases)~date
plot(f, data=d, ylab="log(cases)", main="Recorded cases")
m <- lm(f, data=d)
abline(m)
r2 <- summary(m)$r.squared
dbl <- 1/sqrt(2)/coef(m)[2]
legend("bottomright", bty="n", legend=c(paste("r^2: ", format(r2, digits=2)), paste("d: ", format(dbl, digits=2))))
```

The parameter d in the plot is the doubling time in days, using the [rule of 70](https://en.wikipedia.org/wiki/Rule_of_72).

We can predict the number of cases at the end of the month and see it is in line with estimates from the government.

```{r}
exp(predict(m, newdata=data.frame(date = as.Date("2020-03-30"))))
```

## Logistic Phase
In the real world, exponentials don't continue indefinitely; after some time, they flatten into a [logistic](https://en.wikipedia.org/wiki/Logistic_function) (sigmoid) [curve](http://www.sci.wsu.edu/math/faculty/hudelson/logisticlesson.html).

```{r}
d$lag.cases = c(NA,head(d$cases, -1))
d$growth = d$cases / d$lag.cases
plot(growth ~ date, data=d)
```

The exponential phase ends when the growth rate slows to 1. (However this is not currently the case: the data is too noisy.)