---
title: "Covid-19 in Ireland"
output: 
  html_document:
    theme: united
params:
  output_dir: "docs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

We perform simple mathematical modelling from government-published data, of recorded cases of Covid-19 in the Republic of Ireland.

For a refresher on exponential growth, please see the 3Blue1Brown [video](https://www.youtube.com/watch?v=Kas0tIxDvrg).

## Reading and Plotting the data

The daily data is available at an Irish Government [website](https://www.gov.ie/en/news/7e0924-latest-updates-on-covid-19-coronavirus/).

```{r message=FALSE}
d <- read.csv(file="~/src/covid/data.csv", header=TRUE, sep=",")
d[is.na(d)] <- 0
d$date <- as.Date(d$date)
d$day <- as.integer(d$date - d$date[1])
d$total_cases <- cumsum(d$cases)
d$total_deaths <- cumsum(d$deaths)
```

From the plot below, it is clear we are currently in the exponential-growth phase.

```{r}
f <- log(total_cases)~date
plot(f, data=d, ylab="log(cases)", main="Recorded cases")
m <- lm(f, data=d)
abline(m)
legend("bottomright", bty="n", legend=paste("r^2: ", format(summary(m)$r.squared, digits=2)))
```

```{r}
exp(predict(m, newdata=data.frame(date = as.Date("2020-03-30"))))
```